Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1}

### 2024-10-07 doc/VerticalArrangements.tex
Gloria's 10/3 edits

\title{Beyond Horizontal and Vertical:\\ Simulating the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson, along with seminar participants at the U.S. Department of Justice and the Federal Trade Commission.}}

@@

\begin{abstract}
\noindent We study simulated mergers that have both horizontal and vertical aspects using a model with Bertrand competition downstream and Nash Bargaining upstream.  We apply this framework to the Republic/Santek merger in waste management services, comparing the actual merger to different counterfactual mergers involving various vertical and horizontal combinations.  Using this example, we find almost no instances where the elimination of double marginalization outweighs merger harms to consumers.  Furthermore, we show how analyzing the upstream and downstream markets separately can lead to inaccurate results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely horizontal or purely vertical.  This is especially true for the mergers involving large firms that attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and policy analysis has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.  We have little guidance for how our merger simulation models and related tools apply to these complex mergers.

The horizontal versus vertical distinction had been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines are from 2010, while the most recent Vertical Merger Guidelines are from 2020.  They have both been superseded by a joint document issued in 2023.  }  However, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical aspects of competition.  As part of the revision effort, these agencies requested public comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''  }  In this paper, we address this question by exploring the implications that a standard merger simulation model has for the welfare effects of complex mergers.

@@

Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market. We highlight how the results are effected by several key parameters and assumptions, namely the market elasticity, the use of flat logit versus nested demand preferences, and the bargaining parameter.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts and misrepresent the extent of consumer harm.

This empirical example allows us to explore the implications of the model for complex mergers.  Starting from the Republic/Santek merger as proposed, we compare our baseline simulation results to those for counterfactual mergers between other firms in the market and to those for counterfactual mergers between different combinations of the upstream and downstream components of Republic and Santek.  This exercise facilitates comparison between how the model behaves for complex mergers versus the more straightforward vertical and horizontal mergers that appear in, for instance, \citeN{ST2021}.  Based on these simulations, we find that the complex mergers we study all result in net harm to consumers, despite the presence of large EDM effects in certain cases.
 Only the counterfactual pure vertical merger between the Santek upstream division and the Republic downstream division provides net benefits to consumers.

There are a few caveats to our analysis.  First, products in our merger simulations do not experience any changes in their qualities or costs, other than movements in their wholesale prices due to shifts in bargaining leverage.  Therefore, we rule out merger efficiencies as they are conceptualized in, for example, \citeN{werden96}, and we instead focus on EDM as the main channel through which mergers may benefit consumers.  Second, we model upstream bargaining as being over linear fees, rather than over two-part tariffs or other compensation schedules.  Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.  Although we make this assumption in part for simplicity, it also may be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.  In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.  The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.  }  Fourth, we rely on logit demand for our baseline results, which limits the ways in which consumers substitute between products.  We also provide sensitivity analyses using the nested logit.

Our paper is related to the literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.  \citeN{Riordan2008} has an excellent summary. Broadly speaking, this research explores how incentives to disadvantage rivals could arise with vertical integration, and in turn shows how vertical mergers could negatively impact consumer welfare.  See, for example, \citeN{SalopScheffman1983}, \citeN{Salinger1988}, \citeN{OSS1990}, and \citeN{Riordan1998}.  In a number of these models, vertically integrated firms raise input prices to competitors or exit the input market entirely.  Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming posted or non-negotiated prices.  However, many of the incentives towards RRC, which usually involve trading off profits from input sales against those from downstream sales, are similar to those that appear in our model. These theory papers tend to focus on pure vertical mergers rather than complex mergers, in order to maintain mathematical tractability.

Meanwhile, the structural empirical industrial organization literature has also studied the competitive effects of vertical integration as part of a greater exploration of the impacts of bargaining power in vertical supply chains. Many of these papers are based on the model developed in \citeN{HW1988}, modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.  These models have been applied to several sectors that are frequent targets of vertical integration, such as healthcare (e.g. \citeN{HL2017}) and video programming distribution (e.g. \citeN{CLWY2017}).  These structural empirical papers explore the welfare effects of various competitive interactions, including complex mergers, in specific industries.\footnote{There is a companion empirical literature that relies more on regression analysis to identify changes in market outcomes due to vertical integration (e.g. \citeN{Chipty2001} on video programming distribution).  }  Our framework is similar to the models appearing in these papers, but it has been simplified for use with limited data, with the goal of being more accessible to practitioners and policymakers.

@@

Turning to the upstream market, profits for wholesaler $w$ are given by
\begin{equation}
\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M,
\label{eq: wh profit}
\end{equation}
where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.  Wholesale prices are set via bilaterial Nash Bargaining between retailers and wholesalers.  The bargaining problem for wholesale price $p^W_{rw}$  takes the form,
\begin{equation}
\max_{p^W_{rw}} (\pi^r - d^r (\mathcal{W}^r \setminus \{w\}))^{\lambda_w} (\pi^w - d^w (\mathcal{R}^w \setminus \{r\}))^{1 - \lambda_w},
\label{eq: bargain}
\end{equation}
where $\lambda_w \in [0,1]$ captures the bargaining power of retailers relative to wholesaler $w$.  The $d^r (\mathcal{W}^r \setminus \{w\})$ is the disagreement payoff for the retailer, and $d^w (\mathcal{R}^w\setminus \{r\})$ is the disagreement payoff for the wholesaler.  The disagreement payoff for the retailer is the profits it would earn if it ceases to sell this wholesale product,
\begin{equation}
d^r (\mathcal{W}^r \setminus \{w\}) = \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}(\mathcal{W}^r \setminus \{w\}) M,
\end{equation}
where $s_{rx}(\mathcal{W}^r \setminus \{w\})$ is the realized downstream market share for product $rx$ when retailer $r$ does not sell wholesaler $w$'s product.  Similarly, the disagreement payoff of the wholesaler is
\begin{equation}
d^w (\mathcal{R}^w \setminus \{r\}) = \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] s_{xw}(\mathcal{W}^r \setminus \{w\}) M,
\end{equation}
which is earned if wholesaler $w$ does not offer its product to retailer $r$.

We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{As explained in \citeN{ST2021}, this assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.  }  Then the first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
\begin{equation}
\begin{split}
&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
&\frac{1-\lambda_w}{\lambda_w} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right).
\end{split}
\label{eq: upstream foc}
\end{equation}
The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) \equiv s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.  Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.

@@

Turning to input prices, when wholesaler $v$ bargains with unaffiliated retailer $j$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
& - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation}
which reflects the change in the disagreement payoff coming from the merger with retailer $t$.  Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $j$, these losses are softened due to a potential for diversion to retailer $t$, which we label the ``RRC effect.''\footnote{The assumption of simultaneous downstream price setting rules out RRC effects in the style of \citeN{SalopScheffman1983}, in that here the wholesaler does not have an incentive to raise prices in order to directly impact the pricing strategies of rival retailers.  Rather, our model has the RRC effect seen in \citeN{CLWY2017}, where the wholesaler has better bargaining leverage over rival retailers due to the possibility of increased sales volume (conditional on prices) for their merged downstream retail partner should negotiations break down.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.  These impacts tend to raise the resulting input price to retailer $j$.

@@

Next consider a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$.  Such a combination has a vertical component and a horizontal component.  The first order condition for setting the downstream price $p_{tv}$ becomes
\begin{equation}
\begin{split}
&\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} + s_{tv} + \overbrace{[p_{tv} - c^W_{tv} - c^R_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}^{\text{direct EDM effect}}+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t, s\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_{\text{upstream UPP effect}} \\ &+\underbrace{\underbrace{[p_{sv} - c^W_{sv} - c^R_{sv}]\frac{\partial s_{sv}}{\partial p_{tv}}}_{\text{indirect EDM Effect}} + \sum_{x \in \mathcal{W}^s \setminus \{v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \frac{\partial s_{sx}} {\partial p_{tv}}}_{\text{downstream UPP effect}}= 0.
\end{split}
\label{eq: intdmerger own ret downstream foc}
\end{equation}
Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $t$ raises its prices.  EDM between $v$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $v$ sold by retailer $s$, those units may earn a larger margin.  This impact, which comes through the interaction of EDM with UPP, we label the ``indirect EDM effect.''

If instead an integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the resulting first order condition for product $tv$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc} with some changes.  The downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of the new partner wholesaler.  The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would tend to raise the value of diverted sales to that product.  Whether these additional incentives to raise prices will dominate the direct EDM impact is an empirical question.  Furthermore, if we instead considered a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.

>>>>

\title{Beyond Horizontal and Vertical:\\ Measuring the Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson, along with seminar participants at the U.S. Department of Justice and the Federal Trade Commission.}}

@@

\begin{abstract}
\noindent We study mergers that have both horizontal and vertical aspects using a simulation model with Bertrand competition downstream and Nash Bargaining upstream.  We apply this framework to the 2021 Republic/Santek merger in waste management services, comparing the actual merger to different counterfactual mergers involving various vertical and horizontal combinations.  Within this example, we find almost no instances where the elimination of double marginalization outweighs merger harms to consumers.  Furthermore, we show how analyzing the upstream and downstream markets separately can lead to inaccurate results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely horizontal or purely vertical.  This is especially true for the mergers involving large firms that attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and policy analysis has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has done little research into ``complex'' mergers, which here we define as mergers that possess both horizontal and vertical aspects.  We have scarce guidance for how our merger simulation models and related tools apply to these complex mergers.

The horizontal versus vertical distinction had been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The most recent Horizontal Merger Guidelines are from 2010, while the most recent Vertical Merger Guidelines are from 2020.  They have both been superseded by a joint document issued in 2023.  }  However, the DOJ and FTC revised their Merger Guidelines in 2023, releasing a single document that discusses both horizontal and vertical competition.  As part of the revision effort, these agencies requested public comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''  }  In this paper, we address this question by exploring the implications that a standard merger simulation model has for the welfare effects of complex mergers.

@@

Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market. We highlight how the results are affected by several key parameters and assumptions, namely the market elasticity, the use of flat logit versus nested demand preferences, and the bargaining parameter.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts and misrepresent the extent of consumer harm.

This empirical example allows us to study the implications of the model for complex mergers.  Starting from the Republic/Santek merger as proposed, we compare our baseline simulation results to those for counterfactual mergers between other firms in the market and to those for counterfactual mergers between different combinations of the upstream and downstream components of Republic and Santek.  This exercise facilitates comparison between how the model behaves for complex mergers versus the more straightforward vertical and horizontal mergers that appear in \citeN{ST2021}.  Based on these simulations, we find that the complex mergers we study all result in net harm to consumers, despite the presence of large EDM effects in certain cases.
 Only the counterfactual pure vertical merger between the Santek upstream division and the Republic downstream division provides net benefits to consumers.

There are a few caveats to our analysis.  First, products in our merger simulations do not experience any changes in their qualities or costs, other than movements in their wholesale prices due to shifts in bargaining leverage.  Therefore, we rule out merger efficiencies as they are conceptualized in, for example, \citeN{werden96}, and we instead focus on EDM as the main channel through which mergers may benefit consumers.  Second, we model upstream bargaining as being over linear fees, rather than over two-part tariffs or other compensation schedules.  Although this assumption means bargaining is inefficient, linear input prices are widely observed in some industries, perhaps because of difficulties with implementing or monitoring more complicated contracts.\footnote{One example of an industry with linear wholesale pricing that has been featured in several vertical merger proceedings is cable programming distribution, as detailed in \citeN{Rogerson2014}.}  Third, our model assumes that upstream and downstream prices are set treating all other prices as fixed.  Although we make this assumption in part for simplicity, it also can be appropriate for industries where prices take time to adjust.\footnote{Examples include situations where products are sold under long-term contracts.  \citeN{CLWY2017} use a similar assumption.  In contrast, \citeN{BBMM2023} allow downstream firms to treat their own prices as adjustable during upstream negotiations.  The authors argue that this setup is useful for settings like grocery stores, where menu costs are low.  }  Fourth, we rely on logit demand for our baseline results, which limits the ways in which consumers substitute between products.  We also provide sensitivity analyses using the nested logit.

Our paper is related to the literature on competition and vertical integration that grew out of a number of theory articles in the 1980s and 1990s.  \citeN{Riordan2008} has an excellent summary. Broadly speaking, this research explores how incentives to disadvantage rivals could arise with vertical integration, and as a corollary shows how vertical mergers could negatively impact consumer welfare.  See, for example, \citeN{SalopScheffman1983}, \citeN{Salinger1988}, \citeN{OSS1990}, and \citeN{Riordan1998}.  In a number of these models, vertically integrated firms raise input prices to competitors or exit the input market entirely.  Unlike in our paper, most of these models are not framed explicitly as bargaining problems, instead assuming posted or non-negotiated prices.  However, many of the incentives towards RRC, which usually involve trading off profits from input sales against those from downstream sales, are similar to those that appear in our model. These theory papers tend to focus on pure vertical mergers rather than complex mergers, in order to maintain mathematical tractability.

Meanwhile, the structural empirical industrial organization literature has studied the competitive effects of vertical integration as part of a greater exploration of bargaining power in vertical supply chains. Many of these papers are based on the model developed in \citeN{HW1988}, modified to allow for empirical estimation, as seen in, for example, \citeN{DKVB2010}.  These models have been applied to several sectors that are frequent targets of vertical integration, such as healthcare (e.g. \citeN{HL2017}) and video programming distribution (e.g. \citeN{CLWY2017}).  These papers explore the welfare effects of various competitive interactions, including complex mergers, in specific industries.\footnote{There is a companion empirical literature that relies more on regression analysis to identify changes in market outcomes due to vertical integration (e.g. \citeN{Chipty2001} on video programming distribution).  }  Our framework is similar to the models appearing in these papers, but it has been simplified for use with limited data, with the goal of being more accessible to practitioners and policymakers.

@@

Turning to the upstream market, profits for wholesaler $w$ are given by
\begin{equation}
\pi^w = \sum_{x \in \mathcal{R}^w} [p^W_{xw} - c^W_{xw}] s_{xw} M,
\label{eq: wh profit}
\end{equation}
where $c^W_{xw}$ is the wholesale marginal cost when dealing with retailer $x$.  Wholesale prices are set via bilaterial Nash Bargaining between retailers and wholesalers.  The bargaining problem for wholesale price $p^W_{rw}$  takes the form,
\begin{equation}
\max_{p^W_{rw}} (\pi^r - d^r (\mathcal{W}^r \setminus \{w\}))^{\lambda_w} (\pi^w - d^w (\mathcal{R}^w \setminus \{r\}))^{1 - \lambda_w},
\label{eq: bargain}
\end{equation}
where $\lambda_w \in [0,1]$ captures the bargaining power of retailers relative to wholesaler $w$.  The $d^r (\mathcal{W}^r \setminus \{w\})$ is the disagreement payoff for retailer $r$, and $d^w (\mathcal{R}^w\setminus \{r\})$ is the disagreement payoff for wholesaler $w$.  The disagreement payoff for the retailer is the profits it would earn if it ceases to sell this wholesale product,
\begin{equation}
d^r (\mathcal{W}^r \setminus \{w\}) = \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] s_{rx}(\mathcal{W}^r \setminus \{w\}) M,
\end{equation}
where $s_{rx}(\mathcal{W}^r \setminus \{w\})$ is the realized downstream market share for product $rx$ when retailer $r$ does not carry wholesaler $w$'s product.  Similarly, the disagreement payoff of the wholesaler is
\begin{equation}
d^w (\mathcal{R}^w \setminus \{r\}) = \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] s_{xw}(\mathcal{W}^r \setminus \{w\}) M,
\end{equation}
which is earned if wholesaler $w$ does not offer its product through retailer $r$.

We assume that the negotiation for a given input price treats other input prices and all downstream prices as given.\footnote{This assumption is equivalent to a situation where all negotiations and choices for downstream prices happen simultaneously.  }  Then the first order condition for the negotiation between retailer $r$ and wholesaler $w$ is
\begin{equation}
\begin{split}
&\overbrace{[p^W_{rw} - c^W_{rw}]s_{rw} - \sum_{x \in \mathcal{R}^w \setminus \{r\}} [p^W_{xw} - c^W_{xw}] \Delta s_{xw}(\mathcal{W}^r \setminus \{w\})}^{\text{wholesaler GFT}} = \\
&\frac{1-\lambda_w}{\lambda_w} \left(\underbrace{[p_{rw} - p^W_{rw} - c^R_{rw}]s_{rw} - \sum_{x \in \mathcal{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathcal{W}^r \setminus \{w\})}_{\text{retailer GFT}}\right).
\end{split}
\label{eq: upstream foc}
\end{equation}
The $\Delta s_{xy}(\mathcal{W}^r \setminus \{w\}) = s_{xy}(\mathcal{W}^r \setminus \{w\}) - s_{xy}$ is the difference in the share of good $y$ sold by retailer $x$ when good $w$ is not offered by retailer $r$ versus when good $w$ is offered by retailer $r$.  Thus, the wholesale price $p_{rw}$ is set such that the payoff to wholesaler $w$ when it sells to retailer $r$ less the payoff when it does not (that is, the gains from trade or ``GFT''), divided by the payoff to retailer $r$ when it buys from wholesaler $w$ less the payoff when it does not, equals the ratio of wholesaler to retailer bargaining power.

@@

Turning to input prices, when wholesaler $v$ bargains with unaffiliated retailer $j$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
& - \overbrace{\overbrace{[p_{tv}-c^W_{tv}-c^R_{tv}] \Delta s_{tv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}] \Delta s_{tx}(\mathcal{W}^j \setminus \{v\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation}
which reflects the change in the disagreement payoff coming from the merger with retailer $t$.  Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $j$, these losses are softened due to a potential for diversion to retailer $t$, which we label the ``RRC effect.''\footnote{The assumption of simultaneous downstream price setting rules out RRC effects in the style of \citeN{SalopScheffman1983}, in that here the wholesaler does not have an incentive to raise prices in order to directly impact the pricing strategies of rival retailers.  Rather, our model has the RRC effect seen in \citeN{CLWY2017}, where the wholesaler has better bargaining leverage over rival retailers due to the possibility of increased sales volume (conditional on prices) for the merged downstream retail partner should negotiations break down.}  Furthermore, the margin on product $v$ sold by retailer $t$ is potentially higher due to EDM than the wholesale margin $v$ would have earned, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.  These impacts tend to raise the resulting input price to retailer $j$.

@@

Next consider a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$.  Such a combination has a vertical component and a horizontal component.  The first order condition for setting the downstream price $p_{tv}$ becomes
\begin{equation}
\begin{split}
&\sum_{x \in \mathcal{W}^t \setminus \{v\}} [p_{tx} - p^W_{tx} - c^R_{tx}]\frac{\partial s_{tx}}{\partial p_{tv}} + s_{tv} + \overbrace{[p_{tv} - c^W_{tv} - c^R_{tv}]\frac{\partial s_{tv}}{\partial p_{tv}}}^{\text{direct EDM effect}}+ \underbrace{\sum_{x \in \mathcal{R}^v \setminus \{t, s\}} [p^W_{xv} - c^W_{xv}] \frac{\partial s_{xv}}{\partial p_{tv}}}_{\text{upstream UPP effect}} \\ &+\underbrace{\underbrace{[p_{sv} - c^W_{sv} - c^R_{sv}]\frac{\partial s_{sv}}{\partial p_{tv}}}_{\text{indirect EDM Effect}} + \sum_{x \in \mathcal{W}^s \setminus \{v\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \frac{\partial s_{sx}} {\partial p_{tv}}}_{\text{downstream UPP effect}}= 0.
\end{split}
\label{eq: intdmerger own ret downstream foc}
\end{equation}
Now there is the possibility for what we call the ``downstream UPP effect'' in the retail market, as the merged firm can recapture sales that are diverted to retailer $s$ when $t$ raises its price.  EDM between $v$ and $s$ can actually increase this UPP impact, because when sales are diverted to product $v$ sold by retailer $s$, those units may earn a larger margin.  This impact, which comes through the interaction of EDM with UPP, we label the ``indirect EDM effect.''

If instead an integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the resulting first order condition for product $tv$ would look similar to equation \eqref{eq: intdmerger own ret downstream foc} with some changes.  The downstream UPP effect would be replaced with an additional upstream UPP effect capturing the value of sales diverted to customers of the new partner wholesaler.  The condition would also include an indirect EDM component, reflecting the ability of retailer $r$ to obtain an additional wholesale product at marginal cost, which in turn would tend to raise the value of diverted sales to that product.  Whether these additional incentives to raise prices dominate the direct EDM impact is an empirical question.  Furthermore, if we instead considered a merger between two integrated retailer/wholesaler pairs, both upstream and downstream UPP effects would enter.

As for wholesale prices, again consider a merger between an integrated retailer/wholesaler $tv$ and a standalone retailer $s$.  When wholesaler $v$ bargains with an unaffiliated retailer $j$ we have the first order condition given by,
\begin{equation}
\begin{split}
&[p^W_{jv} - c^W_{jv}]s_{jv} - \sum_{x \in \mathcal{R}^v \setminus \{t,s,j\}} [p^W_{xv} - c^W_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})\\
& - \overbrace{\sum_{x \in \{t,s\}}\left(\overbrace{[p_{xv}-c^W_{xv}-c^R_{xv}] \Delta s_{xv}(\mathcal{W}^j \setminus \{v\})}^{\text{indirect EDM effect}} - \sum_{y \in \mathcal{W}^x \setminus \{v\}} [p_{xy} - p^W_{xy} - c^R_{xy}] \Delta s_{xy}(\mathcal{W}^j \setminus \{v\})\right)}^{\text{RRC effect}}=\\
&\frac{1-\lambda_v}{\lambda_v} \left([p_{jv} - p^W_{jv} - c^R_{jv}]s_{jv} - \sum_{x \in \mathcal{W}^j \setminus \{v\}} [p_{jx} - p^W_{jx} - c^R_{jx}] \Delta s_{jx}(\mathcal{W}^j \setminus \{v\})\right).
\end{split}
\label{eq: intdmerger wh upstream foc}
\end{equation}
The RRC effect is augmented with the profits emanating from the sales of retailer $s$, in addition to the sales of retailer $t$, both of which may potentially recapture sales should retailer $j$ lose access to product $v$.  The merged firm has additional bargaining leverage as a result.  If instead the integrated retailer/wholesaler $tv$ were to merge with a standalone wholesaler, then the profits earned by firm $tv$ should the negotiation with retailer $j$ fail are augmented with the earnings of the additional wholesaler rather than with the earnings of retailer $s$.  This adds a term similar to the wholesale recapture leverage effect seen in equation \eqref{eq: vmerger ret upstream foc} to the left-hand side of the bargaining first order condition.\footnote{For simplicity, we assume that when a retailer fails to reach an agreement with one wholesaler, that retailer's contracts with all other wholesalers remain in place.  }    In the case of a merger between two integrated retailer/wholesaler firms, all of these effects would appear.
